---
title: "Models III: Linear Models in Ag-Data Science"
author: "Dr. Adrian Correndo"
date: "2025-03-05"
categories: [linear models, R, statistics, agriculture]
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 4
    number-sections: true
    table-class: "table table-striped table-hover"
editor: source
execute:
  echo: true
  warning: false
  message: false
smooth-scroll: true
---

## Introduction

Linear models are foundational in statistical analysis, particularly in agricultural data science. These models allow researchers to evaluate relationships between variables and assess treatment effects in experiments. This document covers the essentials of linear modeling in R using `stats`, `car`, `broom`, `emmeans`, `multcomp`, and `cld` for statistical inference and means comparisons.

## What is a Linear Model?

A **linear model** is a mathematical equation describing the relationship between a response variable (dependent) and one or more explanatory variables (independent). The simplest form is:

$$ Y = \beta_0 + \beta_1 X + ... + \beta_2 X ... + \epsilon $$

where:

$$Y$$ is the dependent variable (response variable), <br/>
$$X$$ is the independent variable (matrix of experimental design), <br/>
$$ \beta_0 $$ is the intercept, <br/>
$$ \beta_1 $$ is the effect of factor #1 on Y (i.e. slope (regression) or mean effect (anova)), <br/>
$$ \beta_2 $$ is the effect of factor #2 on Y (i.e. slope (regression) or mean effect (anova)), <br/>
$$ \epsilon $$ represents error (unexplained variation). <br/>

Note: A polynomial term–a quadratic (squared or $2^{nd}$ order) or cubic ($3^{rd}$ order) term turns a linear regression model into a curve. But because it is X that is squared or cubed, not the Beta coefficient ($ \beta $), it still qualifies as a linear model.

### Regression vs. ANOVA

-   **Regression Analysis**: Used when the explanatory variable(s) are continuous (e.g., predicting yield from nitrogen levels).

-   **ANOVA (Analysis of Variance)**: Used when at least one explanatory variable is categorical (e.g., comparing mean yields across different treatments).

-   Technically, ANOVA is a type of linear model, but its focus is on comparing means across groups (made from categorical predictors), while regression aims to quantify the relationship between continuous variables.

We will use the `data_corn` dataset from the `agridat` package.

```{r}
# Load required packages
library(pacman)
p_load(dplyr, tidyr)
p_load(agridat)
p_load(broom)
p_load(emmeans)
p_load(multcomp, multcompView)
p_load(car) # for assumption checks
p_load(performance) # for assumption checks
```

**Load dataset:**

```{r}
data_corn <- agridat::lasrosas.corn

# Inspect dataset
glimpse(data_corn)
```

## Model Fitting: Linear Regression & ANOVA

### Simple Linear Regression (Continuous Predictor)

```{r}
reg_fit <- lm(yield ~ 1 + nitro, data = data_corn)
summary(reg_fit)

# Comparing analysis of variance options
anova(reg_fit)
car::Anova(reg_fit, type = 2)
car::Anova(reg_fit, type = 3)
```

### Linear Regression, no-intercept

```{r}
reg_noint <- lm(yield ~ -1 + nitro, data = data_corn)
summary(reg_fit)
summary(reg_noint)

# Comparing analysis of variance options
anova(reg_noint)
Anova(reg_noint, type = 3)
Anova(reg_noint, type = 2)
```

### ANOVA (Categorical Predictors)

```{r}
anova_crd <- lm(yield ~ nf, data = data_corn) # assuming CRD
anova_fit <- lm(yield ~ nf + rep, data = data_corn) # assuming RCBD
anova_03 <- lm(yield ~ nf*rep, data = data_corn) # assuming RCBD

Anova(anova_crd, type = 3)
#Anova(anova_fit, type = 3)
Anova(anova_03, type = 3)

```

### ANOVA, no-intercept

```{r}
anova_noint <- lm(yield ~ -1 + nf + rep, data = data_corn)
Anova(anova_noint, type = 3)

summary(anova_fit)
summary(anova_noint)
```

## Model Assumptions

Linear models assume:

1.  **Linearity**: Relationship between predictors and response is linear (continuous).

```{r}
# Residual diagnostics
par(mfrow=c(2,2))
plot(anova_fit)
```

2.  **Normality of Residuals**: Residuals should be normally distributed.

```{r}
# Normality test
shapiro.test(resid(anova_fit))
```

3.  **Homoscedasticity**: Equal variance across all levels of predictors.

```{r eval=FALSE}
# Homoscedasticity check
leveneTest(anova_crd) #1-way anova, only nf and CRD
leveneTest(anova_fit) #2-way anova including blocks
```

4.  **Independence**: Observations are independent (i.e. the “error” of replications is independent).

There is no test for independence. You have to make sure you specify the error-structure correctly for potential autocorrelation (e.g. blocks, split-plots, repeated measures, etc.).

5.  **Performance package**
With the `performance` package, we could check all at once.
```{r fig.width = 8, fig.height = 9}
performance::check_model(anova_fit)
```

## Model Selection: AIC/BIC Criteria

### Candidate models
These are all fixed-effect models.
```{r}
# null model
lm_00 <- lm(yield ~ 1, data = data_corn)
# Simplest model
lm_01 <- lm(yield ~ nf + rep, data = data_corn)
# Add year
lm_02 <- lm(yield ~ nf + year + rep, data = data_corn)
# Add topo
lm_03 <- lm(yield ~ nf + topo + rep, data = data_corn)
# Add year and topo
lm_04 <- lm(yield ~ nf + year + topo + rep, data = data_corn)

# Main effects and interactions
lm_05 <- lm(yield ~ nf*year*topo + rep, data = data_corn)
#lm_05 <- lm(yield ~ nf + year + topo + nf:year + nf:topo + year:topo + nf:year:topo + rep, data = data_corn)

```

### Selection criteria
#### F-Test
- **F-tests** evaluate if added predictors significantly improve model fit via sum of squares and degrees of freedom.
```{r}
anova(lm_01, lm_02, lm_03, lm_04, lm_05)
```

#### AIC (Akaike Information Criterion)
- **AIC** balances goodness of fit with model complexity.
```{r}
AIC(lm_01, lm_02, lm_03, lm_04, lm_05)
```

#### BIC (Bayesian Information criterion)
- **BIC** applies a stricter penalty for complexity, favoring simpler models.
```{r}
BIC(lm_01, lm_02, lm_03, lm_04, lm_05)
```

## Significance of effects

```{r}
# Compare Anova sum of squares
Anova(lm_05, type = 2)
Anova(lm_05, type = 3)
```

## Means Comparisons with `emmeans` and `cld`
### Interaction
```{r}
# Pairwise comparisons among treatment means
emmeans(lm_05, pairwise ~ year:topo) %>% 
  cld(., level = 0.05, decreasing = F)
```
### Alternative 
Using the same model, same sum of squares we could group comparisons differently. In this case, showing comparisons of topo means, grouped by "year"
```{r}
emmeans(lm_05, pairwise ~ topo, by = "year") %>% 
  cld(., level = 0.05, decreasing = FALSE, Letters = letters) # add letters

# By default in R
LETTERS
letters
```

## Interpreting Coefficients
### Regression
-   **Intercept ($\beta_0$)**: Baseline value of the dependent variable. Is the value of Y, when X = 0.

-   **Slope ($\beta_1$)**: Change in response variable (Y) per unit increase in predictor (X).

-   **p-value**: Significance of predictor effect.

We can extract regression coefficient estimates with the 'coef()' function, or with the 'tidy()' function of the "broom" package.
```{r}
# Extracting coefficients with coef()
coef(reg_fit)

# Tidy summary of coefficients
reg_coefs <- broom::tidy(reg_fit)
reg_coefs
```

### ANOVA
-   **Intercept**: the reference or benchmark level (baseline mean) for categorical predictors.
-   **Factor Levels**: estimates of mean differences "with respect to" the intercept (if any, if not, from zero).
-   **p-value**: whether a factor significantly differs from baseline.

We can also extract ANOVA coefficient estimates with the 'coef()' function, or with the 'tidy()' function of the "broom" package.
```{r}
# Extracting coefficients with coef()
coef(lm_05)

# Tidy summary of coefficients
anova_coefs <- broom::tidy(lm_05)
anova_coefs

performance::r2(lm_00)
performance::r2(lm_01)
performance::r2(lm_02)
performance::r2(lm_03)
performance::r2(lm_04)
performance::r2(lm_05)
```






## Conclusion

Linear models are essential for agricultural research, helping to quantify relationships and test hypotheses. This quick guide covered essentials for regression (continuous predictors), ANOVA (categorical predictors), assumption checks, model selection, and means comparisons using `emmeans`, and coefficients' extraction with `broom`.
